---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(boot)
library(broom.mixed)
library(collapse)
library(dataverse)
library(gutenbergr)
library(janitor)
library(knitr)
library(marginaleffects)
library(modelsummary)
library(rstanarm)
library(tidybayes)
library(tidyverse)
df <- read.csv(here::here("data/02-analysis_data/analysis_data.csv"))

```

```{r}
#| warning: false
#| message: false
filtered_df <- df |> filter(year==2021) |> filter(age!=996) |> filter(marstat!=99) |> filter(!educ %in% c(996,997,998,999))
labeled_df <- filtered_df |> mutate(
  label_outcome = if_else(outcome==1, "Yes", "No"),
  label_outcome = as_factor(label_outcome),
  label_sex = if_else(sex==1,"Male", "Female"),
  label_education = case_when(
      educ==100~"Grade 12 or less, no high school diploma or equivalent",
      educ==101~"Grade 8 or less (no further detail)",
      educ==102~"Never attended/kindergarten only",
      educ==103~"Grades 1-11 (no further detail)",
      educ==104~"Grade 1",
      educ==105~"Grade 2",
      educ==106~"Grade 3",
      educ==107~"Grade 4",
      educ==108~"Grade 5",
      educ==109~"Grade 6",
      educ==110~"Grade 7",
      educ==111~"Grade 8",
      educ==112~"Grade 9-12, no diploma (no further detail)",
      educ==113~"Grade 9",
      educ==114~"Grade 10",
      educ==115~"Grade 11",
      educ==116~"12th grade, no diploma",
      educ==200~"High school diploma or GED",
      educ==201~"High school graduate",
      educ==202~"GED or equivalent",
      educ==300~"Some college, no 4yr degree",
      educ==301~"Some college, no degree",
      educ==302~"AA degree: technical/vocational/occupational",
      educ==303~"AA degree: academic program",
      educ==400~"Bachelor's degree (BA,AB,BS,BBA)",
      educ==500~"Master's, Professional, or Doctoral Degree",
      educ==501~"Master's degree (MA,MS,Med,MBA)",
      educ==502~"Professional (MD,DDS,DVM,JD)",
      educ==503~"Doctoral degree (PhD, EdD)",
      educ==504~"Other degree",
      educ==505~"Professional School or Doctoral degree, topcoded (MD, DDS, DVM, JD, PhD, EdD)"
  ),
  label_education = factor(
    label_education,
    levels=c(
      "Grade 12 or less, no high school diploma or equivalent",
      "Grade 8 or less (no further detail)",
      "Never attended/kindergarten only",
      "Grades 1-11 (no further detail)",
      "Grade 1",
      "Grade 2",
      "Grade 3",
      "Grade 4",
      "Grade 5",
      "Grade 6",
      "Grade 7",
      "Grade 8",
      "Grade 9-12, no diploma (no further detail)",
      "Grade 9",
      "Grade 10",
      "Grade 11",
      "12th grade, no diploma",
      "High school diploma or GED",
      "High school graduate",
      "GED or equivalent",
      "Some college, no 4yr degree",
      "Some college, no degree",
      "AA degree: technical/vocational/occupational",
      "AA degree: academic program",
      "Bachelor's degree (BA,AB,BS,BBA)",
      "Master's, Professional, or Doctoral Degree",
      "Master's degree (MA,MS,Med,MBA)",
      "Professional (MD,DDS,DVM,JD)",
      "Doctoral degree (PhD, EdD)",
      "Other degree",
      "Professional School or Doctoral degree, topcoded (MD, DDS, DVM, JD, PhD, EdD)"
    )
  )
)

labeled_df |>
  ggplot(aes(x = label_education, fill = label_outcome)) +
  stat_count(position = "dodge") +
  facet_wrap(facets = vars(label_sex)) +
  theme_minimal() +
  labs(
    x = "Highest education",
    y = "Number of respondents",
    fill = "Mental Depression"
  ) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "bottom")
```

```{r}
#| warning: false
#| message: false
labeled_df_reduced <- labeled_df |> slice_sample(n = 1000)
mental_illness <-
  stan_glm(
    label_outcome ~ label_sex + label_education,
    data = labeled_df_reduced,
    family = binomial(link = "logit"),
    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
    prior_intercept = 
      normal(location = 0, scale = 2.5, autoscale = TRUE),
    seed = 853
  )
saveRDS(
  mental_illness,
  file = here::here("models/mental_illness.rds")
)
```
```{r}
#| warning: false
#| message: false
mental_illness <- readRDS(file = here::here("models/mental_illness.rds"))

modelsummary(
  list(
    "Mental Illness Exist" = mental_illness
  ),
  statistic = "mad"
  )
```
# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....






# Data {#sec-data}

## Overview

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

Overview text

## Measurement
	
Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.



Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it. 

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.








# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.  

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.


### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.


# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```




# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows... 

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```



\newpage


# References


